---
title: "Weeks 5 and 6"
output: html_notebook
---

# Set Up

```{r Packages}
#Instal packages
install.packages("tidyverse")
install.packages("janitor")
install.packages("haven")
install.packages("dplyr")
install.packages("labelled")

#Load packages
library(tidyverse)
library(janitor)
library(haven)
library(dplyr)
library(labelled)

```

# Prepare Data

```{r Demo Data}

# Read Demo data for 2015 - 2016 Cycle
demo15_16 <- read_xpt("../Data/DEMO_I.XPT")%>%
  clean_names()%>%
  select(seqn, riagendr, ridageyr, ridreth1)  %>%
  rename(gender=riagendr,
         age = ridageyr,
         race = ridreth1) %>%
  mutate(race = factor(race, levels = c(1:5), 
                       labels = c("mexican american", "other hispanic", 
                                  "non-hispanic white","non-hispanic black",
                                  "other race - including multi-racial")))  %>% 
  mutate(gender=factor(gender, levels = c(1, 2), labels = c("male", "female")))  %>%
  set_variable_labels(gender='Gender',
                      race='Race/Hispanic origin')

# Read Demo data for 2017 - 2018 Cycle
demo17_18 <- read_xpt("../Data/DEMO_I.XPT") %>%
  clean_names() %>%
  select(seqn, riagendr, ridageyr, ridreth1)%>%
  rename(gender=riagendr,
         age = ridageyr,
         race = ridreth1) %>%
  mutate(race = factor(race, levels = c(1:5), 
                       labels = c("mexican american", "other hispanic", 
                                  "non-hispanic white","non-hispanic black",
                                  "other race - including multi-racial"))) %>% 
  mutate(gender=factor(gender, levels = c(1, 2), labels = c("male", "female"))) %>%
  set_variable_labels(gender='Gender',
                      race='Race/Hispanic origin')


# Combine Demo
demo_all <- union(demo15_16, demo17_18)

```


```{r BP}

# Read BP data for 2015 - 2016 Cycle
bpc15_16 <- read_xpt("../Data/BPQ_I.XPT")%>% 
  clean_names()%>%
  select(seqn, bpq020)%>%
  rename(told_had_bp=bpq020)%>% 
  mutate(told_had_bp=factor(told_had_bp, levels = c(1, 2), 
                            labels = c("yes", "no"))) %>%
  set_variable_labels(told_had_bp='Ever told you had high blood pressure')

# Read BP data for 2017 - 2018 Cycle

bpc17_18 <- read_xpt("../Data/BPQ_J.XPT")%>% 
  clean_names()%>%
  select(seqn, bpq020)%>%
  rename(told_had_bp=bpq020)%>% 
  mutate(told_had_bp=factor(told_had_bp, levels = c(1, 2), 
                            labels = c("yes", "no")))%>%
  set_variable_labels(told_had_bp='Ever told you had high blood pressure')

# Combine BP data 

bpc_all <- union(bpc15_16, bpc17_18)

```

```{r BMI }

# Read BMI data for 2015 - 2016 Cycle

bmi15_16 <- read_xpt("../Data/BMX_I.XPT") %>%
  clean_names()%>%
  select(seqn, bmxwt, bmxht) %>%
  rename(weight = bmxwt,
         height = bmxht) %>%
  mutate(bmi=weight/(height/100)^2)   %>% 
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "underweight",
    bmi >= 18.5 & bmi < 24.9 ~ "normal weight",
    bmi >= 25 & bmi < 29.9 ~ "overweight",
    bmi >= 30 ~ "obese"
  ))  %>%
  set_variable_labels(bmi='Body mass index',
                      bmi_category='Categorical body mass index')


# Read BMI data for 2017 - 2018 Cycle
bmi17_18 <- read_xpt("../Data/BMX_J.XPT")  %>%
  clean_names() %>%
  select(seqn, bmxwt, bmxht) %>%
  rename(weight = bmxwt,
         height = bmxht)  %>%
  mutate(bmi=weight/(height/100)^2)  %>% 
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "underweight",
    bmi >= 18.5 & bmi < 24.9 ~ "normal weight",
    bmi >= 25 & bmi < 29.9 ~ "overweight",
    bmi >= 30 ~ "obese"
  )) %>%
  set_variable_labels(bmi='Body mass index',
                      bmi_category='Categorical body mass index')


# Combine BMI data
bmi_all <- union(bmi15_16, bmi17_18)

```


```{r Combine All}
# Combine all data
all <- demo_all %>%
  left_join(bpc_all, by = "seqn") %>%
  left_join(bmi_all, by = "seqn")
```

# Week 5 Tasks

## Task 1: Descriptive Statistics

```{r BMI and Age}

# Calculate descriptive statistics
descr_stat <- all %>%
  summarise(
    mean_bmi = mean(bmi, na.rm = T),
    sd_bmi = sd(bmi, na.rm = T),
    iqr_bmi = IQR(bmi, na.rm = T),
    mean_age = mean(age, na.rm = T),
    sd_age = sd(age, na.rm = T),
    iqr_age = IQR(age, na.rm = T)
  )

# print
print(descr_stat)
```

```{r  BMI and Age (By Gender)}

# Calculate descriptive statistics
descr_stat_gender <- all %>%
  group_by(gender) %>%
  summarise(
    mean_bmi = mean(bmi, na.rm = T),
    sd_bmi = sd(bmi, na.rm = T),
    iqr_bmi = IQR(bmi, na.rm = T),
    mean_age = mean(age, na.rm = T),
    sd_age = sd(age, na.rm = T),
    iqr_age = IQR(age, na.rm = T)
  )

# print
print(descr_stat_gender)

```
## Task 2: Hypothesis Testing

H0: No difference in mean BMI between males and females.
H1: There is difference in mean BMI between males and females.

```{r T-test for Mean BMI Comparison by Gender}

# Perform t-test to compare mean BMI between males and females
ttest_result <- t.test(bmi ~ gender, data = all, na.rm = T)
# Print the t-test result
print(ttest_result)

```
We reject null hypothesis as p-value is < 0.05.



# ANOVA

H0: No difference in mean age across BMI categories.
H1: There is difference in mean age across BMI categories.


```{r ANOVA for Mean Age Across BMI Categories}
# NOVA to compare mean age across BMI categories
anova_result <- aov(age ~ bmi_category, data = all) 

# Print the ANOVA result
anova_summary <- summary(anova_result)

# Perform adhoc if significant 

# Extract p-value from ANOVA summary directly
anova_p_value <- anova_summary[[1]]$`Pr(>F)`[1] # Extracting the p-value
if (anova_p_value < 0.05) { 
 post_hoc <- TukeyHSD(anova_result)
 print(post_hoc)
} else {
 cat("ANOVA is not statistically significant; no need for post-hoc analysis.\n")
}

```

P value is <0.05 across all categories. There is a statistical diffirence.


# CHI

H0: There is an association between BMI cateogry and BP diagnosis.
H1: There is no association between BMI cateogry and BP diagnosis.


```{r Chi for BMI and BP}
# Contingency table for BMI category and blood pressure

contg_tb <- table(all$bmi_category, all$told_had_bp)

# Perform Chi-Square test
chi_square_result <- chisq.test(contg_tb)

# Print the Chi-Square test result
print(contg_tb)
```

P value <0.05 so we reject the null hypothesis. There is an association between BMI cateogry and BP diagnosis.



# Week 6 Tasks



## Task 1: Basic Plot Creation


```{r Simple Scatter Plot}
ggplot(data = all, aes(x = bmi, y = weight)) +
 geom_point()

```


```{r Customised Scatter Plot}
ggplot(data = all, aes(x = bmi, y = weight, colour = age)) +
  geom_point(size = 3, shape = 16, alpha = 0.8) +
  scale_colour_gradient(low = "skyblue", high = "darkblue") +
  labs(title = "BMI vs Weight",
       subtitle = "Coloured by Age",
       x = "Body Mass Index (BMI)",
       y = "Weight (kg)",
       colour = "Age") +
  theme_minimal(base_size = 15) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title = element_text(face = "bold"))
```

```{r Categorising age}

all <- all %>%
  mutate(age_category = case_when(
    age < 18 ~ "Under 18",
    age >= 18 & age < 30 ~ "18-29",
    age >= 30 & age < 45 ~ "30-44",
    age >= 45 & age < 60 ~ "45-59",
    age >= 60 ~ "60+"
  ))
```


```{r Simple Bar chart}
ggplot(data = all %>% filter(age_category != "Under 18"), aes(x =
age_category, fill = told_had_bp)) +
 geom_bar(position = "dodge")

```


```{r Cleaned Bar Chart}
ggplot(data = all %>% filter(age_category != "Under 18"),
       aes(x = age_category, fill = told_had_bp)) +
  geom_bar(position = "dodge") +
  labs(title = "Hypertension Status by Age Category",
       x = "Age Category",
       y = "Number of Individuals",
       fill = "Hypertension Status") +
  theme_minimal(base_size = 15) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"))

```

## Task 2: Applying Layers

```{r Facets}
ggplot(data = all, aes(x = bmi, y = weight)) +
 geom_point() +
 facet_grid(gender ~ told_had_bp)

```

```{r Stat Transformation}
ggplot(data = all, aes(x = bmi, y = weight, color = gender)) +
 geom_point() +
 stat_smooth(method = "lm", se = TRUE) +
 theme_minimal()

```

```{r Coordinate Systems}
ggplot(data = all, aes(x = bmi, y = weight, color=gender)) +
 geom_point() +
 coord_cartesian(xlim = c(15, 40), ylim = c(40, 150))
```

## Task 3: Customizing Themes and Labels
```{r Themes}

ggplot(data = all, aes(x = bmi, y = weight)) +
 geom_point() +
 theme_classic()

```

```{r Labels}
ggplot(data = all, aes(x = bmi, y = weight, color = gender)) +
 geom_point() +
 labs(
 title = "Relationship Between BMI and Weight",
 x = "BMI",
 y = "Weight (kg)",
 color = "Gender"
 ) +
 theme(
 plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
 axis.title.x = element_text(size = 12),
 axis.title.y = element_text(size = 12)
 )
```
## Task 4: Save Your Plots

```{r}
ggsave("../Output/bmi_weight_scatter.png", width = 8, height = 6, dpi = 300)

```
