---
title: "Week 10 R Exercise"
output:
  pdf_document: default
  html_notebook: default
---



## Load Libraries

```{r}
library(tidyverse)
library(janitor)
library(haven)
library(dplyr)
library(labelled)
library(ggplot2)

```

## Read Data

```{r}
all <- readRDS("../Data/all_data_updated.rds")

```

# Task 1: Building a Simple Logistic Regression Model

```{r}
# Filter out "Under 18" age group 
filtered_data <- all %>% filter(age_category != "Under 18") 

 

# Ensure age_cat is a factor without unused levels 
# Convert to factor
filtered_data$age_category <- factor(filtered_data$age_category) 
filtered_data$age_category <- droplevels(filtered_data$age_category) 

# Create a binary variable for smoking status
filtered_data <- filtered_data %>%
  mutate(smoking01 = ifelse(smoking_status == "current", 1, 0))

# Fit the logistic regression model 
model <- glm(smoking01 ~ age_category, data = filtered_data, family = "binomial") 

 

# Print model summary 
summary(model) 
```
### Intercept (-1.58605):
The intercept represents the log-odds of being a current smoker for the reference age group (18-29 years).

Odds of smoking for 18-29 group=exp(−1.58605)≈0.204
This means that the odds of being a current smoker in the 18-29 age group are approximately 0.2 to 1 (i.e., about 20%)


### Age Category Effects:
The coefficients for the age_category variable represent the difference in log-odds of being a current smoker compared to the reference group (18-29).

#### Age Group 30-44 (0.31093):
The log-odds of being a current smoker are higher by 0.31 compared to the 18-29 group.
Converting to odds ratio: exp(0.31093)≈1.364
This means the odds of being a current smoker are about 36% higher in the 30-44 age group compared to the 18-29 group.
p=0.002: Statistically significant increase in odds of smoking compared to the 18-29 group.

#### Age Group 45-59 (0.34207):
The log-odds of being a current smoker are higher by 0.34 compared to the 18-29 group.
Odds ratio:exp(0.34207)≈1.408
The odds of being a current smoker are about 41% higher in the 45-59 group compared to the 18-29 group.
p=0.0008: Statistically significant increase in odds of smoking compared to the 18-29 group.

#### Age Group 60+ (-0.21783):
The log-odds of being a current smoker are lower by 0.22 compared to the 18-29 group.
Odds ratio:exp(−0.21783)≈0.804
This means the odds of being a current smoker are about 20% lower in the 60+ group compared to the 18-29 group.
p=0.036: Statistically significant decrease in odds of smoking compared to the 18-29 group.


```{r Visualising the relationship}
# Predict probabilities for each age category
predicted_probs <- filtered_data %>%
  group_by(age_category) %>%
  summarize(
    smoking_rate = mean(smoking01, na.rm = TRUE)
  )

# Plot the smoking rates by age category
ggplot(predicted_probs, aes(x = age_category, y = smoking_rate)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_text(aes(label = scales::percent(smoking_rate, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  labs(
    title = "Smoking Rates by Age Category",
    x = "Age Category",
    y = "Smoking Rate (Proportion)"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal()
```



# Task 2: Calculating and Interpreting Odds Ratios with Confidence Intervals 

```{r}
# Calculate odds ratios and 95% confidence intervals 
odds_ratios <- exp(coef(model)) 

conf_int <- exp(confint(model)) 

 

# Create a table for reporting 
results <- data.frame( 
  Odds_Ratio = odds_ratios, 
  CI_Lower = conf_int[, 1], 
  CI_Upper = conf_int[, 2] 
) 


print(results) 
```

Intercept:
The confidence interval does not include 1, indicating statistical significance.

Age Group 30-44:
The confidence interval does not include 1, so the result is statistically significant.

Age Group 45-59:
The confidence interval does not include 1, so this result is statistically significant.

Age Group 60+:
The confidence interval does not include 1, so this result is statistically significant.

# Task 3: Creating a Table of Log-Odds, Odds, and Probabilities 

```{r}
# Extract model coefficients 
intercept <- coef(model)[1]  # Intercept term 
coefficients <- coef(model)[-1]  # Coefficients for age categories 

 

# Calculate log-odds, odds, and probabilities for each age category 
log_odds <- c(intercept, intercept + coefficients)  # Log-odds for each age category 

odds <- exp(log_odds)  # Convert log-odds to odds 

probabilities <- odds / (1 + odds)  # Convert odds to probabilities 

 

# Create a data frame with age category names and calculated values 

results <- data.frame( 
  Log_Odds = log_odds, 
  Odds = odds, 
  Probability = probabilities 
) 

 
print(results) 
```

# Task 4: Adding a Categorical Predictor 

```{r}
# Fit a logistic regression model with a categorical predictor 
model_gender <- glm(smoking01 ~ age_category + gender, data = filtered_data, family = "binomial") 

 

# Print model summary 
summary(model_gender) 

 

# Calculate odds ratios and 95% confidence intervals 
odds_ratios_gender <- exp(coef(model_gender)) 
conf_int_gender <- exp(confint(model_gender)) 

 

# Create a table for reporting 
results_gender <- data.frame( 
  Odds_Ratio = odds_ratios_gender, 
  CI_Lower = conf_int_gender[, 1], 
  CI_Upper = conf_int_gender[, 2] 
) 


print(results_gender) 
```

Compared to the reference category (males aged 18-29):
Males aged 30-44 and 45-59 are 38% and 41.3% more likely to smoke, respectively.
Males aged 60+ are 20.4% less likely to smoke compared to the reference group.
Females are 44% less likely to smoke compared to males in the same age category.
